---
title: "Hierarchical modelling"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Theo Rashid"
date: "August 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r packages}
library(tidyverse)
library(nimble)

set.seed(2)
```

This task will involve taking the England or Italy data and fitting a really basic hierarchical model.
It will follow your basic:

- Full pooling
- No pooling
- Partial pooling

Aka radon model example. https://www.tensorflow.org/probability/examples/Multilevel_Modeling_Primer
The difference being our subgroups will be regions of England or Italy.

Load in the data at the level
Estimate the national prevalence/whatever the outcome is

Estimate no pooling
Estimate full pooling

Partial pooling model
Discuss the advantages
Try out different priors for sigma and see how that changes the fit
Run some prior predictive checks for good practice
Get used to diagnosing fit using the posterior and bayesplot package
