---
title: "Introduction to Bayesian Methods"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Robbie M. Parks"
date: "August 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Load pacakges
```{r}
library(here)
library(tidyverse)
library(nimble)
```

## Goal of this computing lab session

This lab will involve taking some concepts from the lectures and introduce you to the way NIMBLE works.

## Introduction to NIMBLE format.

NIMBLE is written in a slightly unusual format if you're used to just using R. It is written in the style of a program called BUGS, which came out a few decades ago and was developed at Imperial College London. 

## First basic examples of NIMBLE and how to use it.

This example will be for a basic regression model using linear predictors (Normal distribution implicitly)

Adapted from https://r-nimble.org/examples

First create some example data for our model
[[[FORMULA HERE IN LaTeX]]]
```{r}
set.seed(1)
p <- 3    # number of explanatory variables
n <- 100   # number of observations
X <- matrix(rnorm(p*n), nrow = n, ncol = p) # explanatory variables
true_betas <- c(c(0.2, 0.5, 0.3)) # coefficients
sigma <- 0.5
y <- rnorm(n, X %*% true_betas, sigma)
```

What does the dataset look like?
```{r}
head(data.frame(y=y,x1=X[,1],x2=X[,2],x3=X[,3]))
```

Create the NIMBLE model
```{r}
code = nimbleCode({
  
  # priors for parameters
  beta0 ~ dnorm(0, sd = 100) # prior for beta0
  beta1 ~ dnorm(0, sd = 100) # prior for beta1
  beta2 ~ dnorm(0, sd = 100) # prior for beta2
  beta3 ~ dnorm(0, sd = 100) # prior for beta3
  sigma ~ dunif(0, 100)      # prior for variance components
  
  # regression formula
  for(i in 1:n) {
    y[i] ~ dnorm(beta0 + beta1*x1[i] + beta2*x2[i] + beta3*x3[i], sd = sigma) # manual entry of linear predictors
  }
  
})
```

Before running NIMBLE, extract data for three predictors and center around zero for better MCMC performance
```{r}
x1 <- X[,1] - mean(X[,1])
x2 <- X[,2] - mean(X[,2])
x3 <- X[,3] - mean(X[,3])
```

Final preparation of data into lists
```{r}
constants <- list(n = n, x1 = x1, x2 = x2, x3 = x3)
data <- list(y = y)
```

Set initial values for MCMC samples
```{r}
inits <- list(beta0 = mean(y), beta1 = 0, beta2 = 0, beta3 = 0, sigma = 1)
```

Create the NIMBLE model from the BUGS code above ready to run in MCMC (redundant)
```{r}
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

The following code will establish which samples will be used in the sampling of the posteriors. If there is a conjugate relationship apparent between prior and posterior (e.g., Normal-Normal, Binomial-Beta, Poisson-Gamma), it will be detected here
```{r}
# mcmcConf <- configureMCMC(model)
```

Specify the number of MCMC samples
```{r}
ni = 10000
```

Run the MCMC simulations
```{r}
t0 = Sys.time()
nimbleMCMC_samples_initial = nimbleMCMC(
                           code = code,
                           data = data,
                           constants = constants, 
                           inits = inits,
                           # model = model,
                           niter = ni,
                           setSeed = 1,
                           summary = TRUE)
t1 =  Sys.time()
t1 - t0
```

What is the summary of each estimated parameter from the samples?
```{r}
nimbleMCMC_samples_initial$summary
```

What do the samples of one of the unknown parameters actually look like? Let's have a look at beta1 (which we know is 0.2)
```{r}
plot(nimbleMCMC_samples_initial$samples[ , 'beta1'], type = 'l', xlab = 'iteration',  
     ylab = expression(Beta_1))
```

So it looks like the samples are converging quickly from the initial parameter to ~0.2. But typically we will throw some samples at the beginning to ensure that the transient samples (which is when the model samples haven't stabilized around a particular value) are not included in calculating estimates of the mean and credible intervals. This is called the 'burn in' or 'warm up period'.

Let's do it again but with a burn in of 1000 samples.

```{r}
ni = 10000
nb = 1000 

t0 = Sys.time()
nimbleMCMC_samples_burnin = nimbleMCMC(code = code,
                           data = data,
                           constants = constants, 
                           inits = inits,
                           # model = model,
                           niter = ni,
                           nburnin = nb,
                           setSeed = 1,
                           summary = TRUE,
)
t1 =  Sys.time()
t1 - t0
```

What is the summary of each estimated parameter from the samples with burn in?
```{r}
nimbleMCMC_samples_burnin$summary
```

Now the samples look to be very tidily centered around 0.2.
```{r}
plot(nimbleMCMC_samples_burnin$samples[ , 'beta1'], type = 'l', xlab = 'iteration',  
     ylab = expression(Beta_1))
```

What about if we want to know which model fits better? Use Watanabe–Akaike information criterion (WAIC)
```{r}

```

What does equivalent frequentist model output look like?
```{r}

```

## Binomial-Beta example.
```{r}

```

## Gamma distribution.
```{r}

```

## Poisson-Gamma example.
```{r}

```
