---
title: "Temporal modelling"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Robbie M. Parks"
date: "August 2023"
output: html_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(tidyverse)
library(nimble)
library(bayesplot)
library(posterior)
library(hrbrthemes)

extrafont::loadfonts()
theme_set(theme_ipsum())

color_scheme_set(scheme = "viridis")

set.seed(2)
```

## Goal of this computing lab session

This lab will involve taking some key temporal modelling concepts, including linear slopes, seasonal terms, random walks, autoregressive models, and forecasting.

## Introduction to data we will be using.

We will be using national death count data for Spain during 2010-2020 as taken from the paper:

V. Kontis, J.E. Bennett, R.M. Parks, T. Rashid, J. Pearson-Stuttard, P. Asaria, B. Zhou, M. Guillot, C.D. Mathers, Y.H. Khang, M. McKee and M. Ezzati. Lessons learned and lessons missed: impact of the coronavirus disease 2019 (COVID-19) pandemic on all-cause mortality in 40 industrialised countries prior to mass vaccination. Wellcome Open Research 2021, 6:279 (paper) (code).

Let's load in the data.
```{r}
data <- read_csv(here("data", "Spain", "data_spain.csv"))
```

Here's a preview of the data
```{r}
data |> head()
```

Now let's summarise the data by week nationally
```{r}
data_national <- data |>
  group_by(week,week_of_year) |>
  summarise(deaths=sum(deaths),
            population=sum(population),
            t2m=mean(t2m),
            weekly_t2m_anomaly=mean(weekly_t2m_anomaly)) |>
  mutate(week=dmy(week)) |>
  arrange(week) |>
  filter(year(week)<2020) # avoiding COVID for now
```

Here's a preview of the summarised national data
```{r}
data_national |> head()
```

Let's plot the number of deaths over time
```{r}
ggplot(data = data_national) + geom_point(aes(x=week,y=deaths)) #+ theme_void()
```

Let's calculate crude death rates over time too
```{r}
data_national = data_national |>
  mutate(rate=100000*deaths/population)
```

Let's plot the death rates over time
```{r}
ggplot(data = data_national) + geom_point(aes(x=week,y=rate))
```

Before running NIMBLE, extract data for week and center around zero for better MCMC performance
```{r}
data_national = data_national |>
  ungroup() |>
  mutate(weekc = row_number()) |>
  mutate(weekc = weekc-mean(weekc)) |>
  mutate(month = month(week))
```

## Linear model over time

The first model we will create and assess is a linear model over time, which assumes that death rates are simply going up or down at a constant rate throughout our study. This is very basic and just to show how to start with some analysis. Since we are dealing with count data, a Poisson model could make sense:

$$
y = max + c
$$

Let's write the NIMBLE code
```{r}
code_linear <- nimbleCode({

  # priors for parameters
  alpha ~ dnorm(0, sd = 100) # prior for alpha
  beta_week ~ dnorm(0, sd = 100) # prior for beta_week

  # regression formula
  for(t in 1:weeks) {
    deaths[t] ~ dpois(mu[t])
    log(mu[t]) <- alpha + beta_week * t + log(population[t])
  }

  # exponentiate the estimated parameters to reverse the log-link function
  alpha_exp <- exp(alpha)
  beta_week_exp <- exp(beta_week)

  # what's the estimated annual rate of change?
  beta_year <- exp(52 * beta_week)
  beta_year_exp <- exp(beta_year)
})
```

Final preparation of data into lists
```{r}
constants <- list(weeks = nrow(data_national))
data <- list(deaths = data_national$deaths,
             week = data_national$weekc,
             population = data_national$population)
```

Set initial values for MCMC samples
```{r}
inits <- list(alpha = 0, beta_week = 0)
parameters_to_monitor = c('alpha','beta_week','alpha_exp','beta_week_exp','beta_year','beta_year_exp')
```

Let's run the model.
```{r}
t0 = Sys.time()
nimbleMCMC_samples_linear <- nimbleMCMC(
  code = code_linear,
  data = data,
  constants = constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 15000,
  nburnin = 3000,
  setSeed = 1,
  samplesAsCodaMCMC = TRUE
)

t1 =  Sys.time()
t1 - t0
```

What is the summary of each estimated parameter from the Poisson model?
```{r}
summarise_draws(nimbleMCMC_samples_linear, default_summary_measures())
summarise_draws(nimbleMCMC_samples_linear, default_convergence_measures())
```

Now the samples look to be very tidily centred.
```{r}
mcmc_trace(nimbleMCMC_samples_linear)
```

Now let's look at the residuals of the fit, which should be randomly distributed around zero without any obvious pattern
```{r}
# TO DO
```

## Linear model over time with random walk term

The residuals look like there may be a seasonal pattern, and so let us attempt to remove that by including a weekly random walk

$$
y = max + c
$$

```{r}
code_weekly_random_walk <- nimbleCode({

  # priors for parameters
  alpha ~ dnorm(0, sd = 100) # prior for alpha
  beta_week ~ dnorm(0, sd = 100) # prior for beta_week
  sigma ~ dunif(0, 100)      # prior for variance of weekly effects

  # regression formula
  for(t in 1:weeks) {

    deaths[t] ~ dpois(mu[t])
    # log(mu[t]) <- alpha + alpha_week[week_of_year[t]] + beta_week * t + log(population[t])
    log(mu[t]) <- alpha + rw[t] + beta_week * t + log(population[t])

  }

  # weekly random effect
  # alpha_week[1] <- 0
  # for (w in 2:n_weeks) {
  #   alpha_week[w] ~ dnorm(0, sigma)
  # }

  # random walk over time
  rw[1] <- 0
  for (t in 2:weeks) {
    rw[t] ~ dnorm(rw[t-1], sigma)
  }

  # exponentiate the estimated parameters to reverse the log-link function
  alpha_exp <- exp(alpha)
  beta_week_exp <- exp(beta_week)

  # what's the estimated annual rate of change?
  beta_year <- exp(52 * beta_week)
  beta_year_exp <- exp(beta_year)
})
```

Final preparation of data into lists
```{r}
constants <- list(weeks = nrow(data_national),
                  n_weeks = max(data_national$week_of_year),
                  week = data_national$weekc,
                  week_of_year = data_national$week_of_year)

data <- list(deaths = data_national$deaths,
             population = data_national$population)
```

Set initial values for MCMC samples
```{r}
inits <- list(alpha = 0, beta_week = 0)
parameters_to_monitor = c('alpha','beta_week','alpha_exp','beta_week_exp','beta_year','beta_year_exp','rw')
```

Let's run the model.
```{r}
t0 = Sys.time()
nimbleMCMC_samples_week_random_walk <- nimbleMCMC(
  code = code_weekly_random_walk,
  data = data,
  constants = constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 100000,
  nburnin = 10000,
  setSeed = 1,
  samplesAsCodaMCMC = TRUE
)

t1 =  Sys.time()
t1 - t0
```

What is the summary of each estimated parameter from the Poisson model?
```{r}
summarise_draws(nimbleMCMC_samples_week_random_walk, default_summary_measures())
summarise_draws(nimbleMCMC_samples_week_random_walk, default_convergence_measures())
```

Now the samples look to be very tidily centred.
```{r}
mcmc_trace(nimbleMCMC_samples_week_random_walk, regex_pars = c('alpha'))
```

What do the random weekly terms looks like?
```{r}
# mcmc_trace(nimbleMCMC_samples_week_random_effect, regex_pars = c('alpha_week'))
```

## Linear model over time with random walk term and overall linear temperature term

Let's see what adding a slope for temperature does
$$
y = max + c
$$

```{r}
code_weekly_random_walk_with_temperature <- nimbleCode({

  # priors for parameters
  alpha ~ dnorm(0, sd = 100) # prior for alpha
  beta_week ~ dnorm(0, sd = 100) # prior for beta_week
  beta_temperature ~ dnorm(0, sd = 100) # prior for beta_temperature
  sigma_week ~ dunif(0, 100) # prior for variance of weekly effects
  sigma_temperature ~ dunif(0, 100) # prior for variance of temperature effects

  # regression formula
  for(t in 1:weeks) {
    deaths[t] ~ dpois(mu[t])
    log(mu[t]) <- alpha +
                  beta_week * t +
                  log(population[t]) + beta_temperature * weekly_t2m_anomaly[t] #+ rw[t]
  }

  # random walk over time
  # rw[1] <- 0
  # for (t in 2:weeks) {
  #   rw[t] ~ dnorm(rw[t-1], sigma_week)
  # }

  # weekly temperature random effect
  # beta_temperature_week[1] <- 0
  # for (w in 2:n_weeks) {
  #   beta_temperature_week[w] ~ dnorm(0, sigma_temperature)
  # }

  # exponentiate the estimated parameters to reverse the log-link function
  alpha_exp <- exp(alpha)
  beta_week_exp <- exp(beta_week)
  beta_temperature_exp <- exp(beta_temperature)

  # what's the estimated annual rate of change?
  beta_year <- exp(52 * beta_week)
  beta_year_exp <- exp(beta_year)
})
```

Final preparation of data into lists
```{r}
constants <- list(weeks = nrow(data_national),
                  n_weeks = max(data_national$week_of_year),
                  week = data_national$weekc,
                  week_of_year = data_national$week_of_year)

data <- list(deaths = data_national$deaths,
             population = data_national$population,
             weekly_t2m_anomaly = round(data_national$weekly_t2m_anomaly,1))
```

Set initial values for MCMC samples
```{r}
inits <- list(alpha = 0, beta_week = 0)
parameters_to_monitor = c('alpha','beta_week','alpha_exp','beta_week_exp','beta_year','beta_year_exp','beta_temperature','beta_temperature_exp')
```

Let's run the model.
```{r}
t0 = Sys.time()
nimbleMCMC_samples_week_random_walk_with_temperature <- nimbleMCMC(
  code = code_weekly_random_walk_with_temperature,
  data = data,
  constants = constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 100000,
  nburnin = 10000,
  setSeed = 1,
  samplesAsCodaMCMC = TRUE
)

t1 =  Sys.time()
t1 - t0
```

What is the summary of each estimated parameter from the Poisson model?
```{r}
summarise_draws(nimbleMCMC_samples_week_random_walk_with_temperature, default_summary_measures())
summarise_draws(nimbleMCMC_samples_week_random_walk_with_temperature, default_convergence_measures())
```

Now the samples look to be very tidily centred.
```{r}
mcmc_trace(nimbleMCMC_samples_week_random_walk_with_temperature)
```

What do the random weekly terms looks like? (Theo)
```{r}

```

## Linear model with random walk over time
```{r}

```

## Linear model with autoregressive term over time
```{r}

```

## Forecasting (what might have happened in 2020 without COVID?)
```{r}

```
