---
title: "Exposure Response Modelling"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Robbie M. Parks, Garyfallos Konstantinoudis"
date: "August 22 2024"
format: html
editor:
  markdown:
    wrap: sentence
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(tidyverse)
library(nimble)
library(sf)
library(posterior)
library(bayesplot)
library(spdep)
library(lubridate)
library(colorspace)
library(coda)

extrafont::loadfonts()
theme_set(hrbrthemes::theme_ipsum())
knitr::opts_chunk$set(fig.align = "center")

set.seed(2)
```

### Goal of this computing lab session

This goal of this lab is to use `NIMBLE` to carry out an exposure-response regression in a variety of ways.

### What's going to happen in this lab session?

During this lab session, we will build:

1.  Linear exposure-response regression;
2.  Piecewise linear exposure-response regression; and
3.  Non-linear exposure-response regression.

## Introduction

We will build on and extend the previous lab the COVID-19 deaths during March-July 2020, in England, at the LTLA geographical level (317 areas), as taken from the published paper:

**Konstantinoudis G**, Padellini T, Bennett JE, Davies B, Ezzati M, Blangiardo M. *Long-term exposure to air-pollution and COVID-19 mortality in England: a hierarchical spatial analysis*.
Environ Int.
2021 Jan:146:106316.
doi: 10.1016/j.envint.2020.106316.
PMID: 33395952 PMCID: PMC7786642

For that analysis, we included 38,573 COVID-19 deaths up to June 30, 2020 at the Lower Layer Super Output Area level in England ($n = 32844$ small areas).
We retrieved averaged NO$_2$ concentration during 2014-2018 from the Pollution Climate Mapping.
We used Bayesian hierarchical models to quantify the effect of air pollution while adjusting for a series of confounding and spatial autocorrelation.

We will build simple Bayesian models to try to understand what is happening in the data.
Once again we will use `NIMBLE` as the basis for our Bayesian model writing.

### Load in data

Let's load in the data

```{r}
data_england <- read_sf(here("data", "England", "COVIDecoregression.shp"))
glimpse(data_england)
summary(data_england)
class(data_england)

# use the code below to simplify the geometry and speed up plotting and other oprtations
data_england_simpler <- data_england #rgeos::gSimplify(as(data_england, "Spatial"), tol = 500)
data_england_simpler <- st_as_sf(data_england_simpler)
data_england_simpler <- cbind(data_england_simpler, data_england %>% mutate(geometry = NULL))
data_england <- data_england_simpler
```

The following code subsets the data to London so the models are quicker to run.
We're going to run the models for London, then load in the samples for England (and pretend we ran for England!).

```{r eval = FALSE}
data_england <- data_england[startsWith(data_england$LTLA, "E09"), ]
ggplot(data = data_england, fill = "NA") +
  geom_sf() +
  theme_void()
```

How many spatial units are in the map?

```{r}
# Obtain the number of LTLAs
n.LTLA <- nrow(data_england)
n.LTLA
```

### Adjacency matrix in R

Convert the polygons to a list of neighbors using the function `poly2nb()`

```{r}
LTLA_nb <- poly2nb(pl = data_england)
LTLA_nb
```

Convert the list you defined previously to `NIMBLE` format (i.e. a list of 3 components adj, num and weights) using the function `nb2WB()` and print a summary of the object.

```{r}
nbWB_A <- nb2WB(nb = LTLA_nb)
names(nbWB_A)
```

## 1. Linear exposure-response regression

Let $\mathcal{D}$ be the observation window of England and $A_1, A_2, \dots, A_N$ a partition denoting the LTLAs in England with $\cup_{i=1}^NA_i = \mathcal{D}$ and $A_i\cap A_j$ for every $i\neq j$.
Let $O_1, O_2, \dots, O_N$ be the observed number of COVID-19 deaths occurred during March-July 2020 in England, $E_1, E_2, \dots, E_N$ is the expected number of COVID-19 deaths and $\lambda_1, \lambda_2, \dots, \lambda_N$ the standardized mortality ratio (recall $\lambda_i = \frac{O_i}{E_i}$).
A standardized mortality ratio of $1.5$ implies that the COVID-19 deaths we observed in the $i$-th area are $1.5$ times higher to what we expected.
Under the Poisson assumption we have:

$$
\begin{equation}
\begin{aligned}
\hbox{O}_i & \sim \hbox{Poisson}(E_i \lambda_i); \;\;\; i=1,...,N\\
\log \lambda_i & = \alpha +  \beta_1 X_{1i} + \beta_2 X_{2i} + \sum_{j=2}^5\beta_{3j} + \theta_i + \phi_i\\
\theta_i &\sim \hbox{Normal}(0, \sigma^2_{\theta_i})\\
{\bf \phi} & \sim \hbox{ICAR}({\bf W}, \sigma_{\phi}^2) \,\, ,  \sum_i \phi_i  = 0 \\
\alpha & \sim \text{Uniform}(-\infty, +\infty) \\
\beta_1, \beta_2 & \sim \mathcal{N}(0, 10) \\
1/\sigma_{\theta}^2 & \sim \hbox{Gamma}(0.5, 0.05) \\
1/\sigma_{\phi}^2 & \sim \hbox{Gamma}(0.5, 0.0005) \\
\end{aligned}
\end{equation}
$$

the terms $\beta_1 X_{1i} + \beta_2 X_{2i} + \sum_{j=2}^5\beta_{3j} X_{3i}$, where $X_{1i}, X_{2i}, X_{3i}$ are the ICU beds, NO$_2$ and IMD in the $i$-th LTLA, $\beta_1, \beta_2, \sum_{j=2}^5\beta_{3j}$ the corresponding effects and $\exp(\beta_1), \exp(\beta_2)$ the relative risk of ICU beds or NO$_2$ for every unit increase and of the ICU beds or NO$_2$.
For instance $\exp(\beta_2) = 1.8$ means that for every unit increase of long term exposure to $NO_2$, the risk (read standardized mortality ratio) of COVID-19 deaths cancer increases by $80\%$.
$exp(\beta_{32}), \beta_{33}, \beta_{34}, \beta_{35}$ are the relative risks compared to the baseline IMD category, ie the most deprived areas.
An $exp(\beta_{35}) = 0.5$ means that the risk of COVID-19 deaths in most affluent areas decreases by $50%$ compared to the most deprived areas.
We will first write the model in `NIMBLE`.

```{r eval=TRUE, echo = TRUE}
BYMecoCode <- nimbleCode({
  # priors
  alpha ~ dflat() # vague prior (Unif(-inf, +inf))
  overallRR <- exp(alpha) # overall RR across study region

  tau.theta ~ dgamma(0.5, 0.05) # prior for the precision hyperparameter
  sigma2.theta <- 1 / tau.theta # variance of unstructured area random effects

  tau.phi ~ dgamma(0.5, 0.0005) # prior on precison of spatial area random effects
  sigma2.phi <- 1 / tau.phi # conditional variance of spatial area random effects

  for (j in 1:K) {
    beta[j] ~ dnorm(0, tau = 1)
    RR.beta[j] <- exp(beta[j])
  }

  RR.beta1_1NO2 <- exp(beta[1] / sd.no2) # get per 1 unit increase in the airpollution (scale back)

  # ICAR prior
  # theta + phi is BYM
  phi[1:N] ~ dcar_normal(adj = adj[1:L], weights = weights[1:L], num = num[1:N], tau = tau.phi, zero_mean = 1)

  # likelihood
  for (i in 1:N) {
    O[i] ~ dpois(mu[i]) # Poisson likelihood for observed counts
    log(mu[i]) <- log(E[i]) + alpha + theta[i] + phi[i] + inprod(beta[], X[i, ])
    # the inprod is equivalent to beta[1]*X1[i] + beta[2]*X2[i] + beta[3]*X32[i] + beta[4]*X33[i] + beta[5]*X34[i] + beta[6]*X35[i]

    SMR[i] <- alpha + theta[i] + phi[i] + inprod(beta[], X[i, ])
    theta[i] ~ dnorm(0, tau = tau.theta) # area-specific RE
    resRR[i] <- exp(theta[i] + phi[i]) # area-specific residual RR
    proba.resRR[i] <- step(resRR[i] - 1) # Posterior probability
  }
})
```

Create data object as required for `NIMBLE`.

```{r eval=TRUE}
n.LTLA <- dim(data_england)[1]

# create the dummy columns for deprivation
data_england <- data_england |>
  mutate(IMD = as_factor(IMD)) |>
  mutate(as.data.frame(model.matrix(~ 0 + IMD, data = pick(everything()))))

# matrix of covariates
Xmat <- cbind(
  scale(data_england$NO2)[, 1],
  scale(data_england$TtlICUB)[, 1],
  data_england$IMD2,
  data_england$IMD3,
  data_england$IMD4,
  data_england$IMD5
)

# Format the data for NIMBLE in a list
covid_data <- list(
  O = data_england$deaths, # observed nb of deaths
  X = Xmat # covariates
)

# number of total covariates
K <- ncol(Xmat)

covid_constants <- list(
  N = n.LTLA, # nb of LTLAs
  K = K, # number of covariates
  sd.no2 = sd(data_england$NO2), # standard deviation of NO2

  # adjacency matrix
  L = length(nbWB_A$weights), # the number of neighboring areas
  E = data_england$expectd, # expected number of deaths
  adj = nbWB_A$adj, # the elements of the neighbouring matrix
  num = nbWB_A$num,
  weights = nbWB_A$weights
)
```

Create the initial values for ALL the unknown parameters:

```{r}
# initialise the unknown parameters, 2 chains
inits <- list(
  list(
    alpha = 0.01,
    beta = rep(0, K),
    tau.theta = 10,
    tau.phi = 1,
    theta = rep(0.01, times = n.LTLA),
    phi = c(rep(0.5, times = n.LTLA))
  ),
  list(
    alpha = 0.5,
    beta = rep(-1, K),
    tau.theta = 1,
    tau.phi = 0.1,
    theta = rep(0.05, times = n.LTLA),
    phi = c(rep(-0.05, times = n.LTLA))
  )
)
```

Which model parameters do you want to monitor?
Set these before running `NIMBLE`.
Call this object `parameters_to_monitor`.

```{r}
parameters_to_monitor <- c("sigma2.theta", "sigma2.phi", "overallRR", "theta", "beta", "RR.beta", "resRR", "proba.resRR", "alpha", "RR.beta1_1NO2")
```

Run the MCMC simulations using the function `nimbleMCMC()`.
If everything is specified reasonably, this runs quickly.

```{r echo=TRUE, eval=FALSE}
tic <- Sys.time()
modelBYMeco.sim <- nimbleMCMC(
  code = BYMecoCode,
  data = covid_data,
  constants = covid_constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 50000,
  nburnin = 30000,
  thin = 10,
  nchains = 2,
  setSeed = 9,
  progressBar = TRUE,
  samplesAsCodaMCMC = TRUE,
  summary = TRUE,
  WAIC = TRUE
)
toc <- Sys.time()
toc - tic
# saveRDS(modelBYMeco.sim, file = "NIMBLE_BYM_LINEAR")
```

```{r}
modelBYMeco.sim <- read_rds(here("labs", "exposure_response", "NIMBLE_BYM_LINEAR"))
```

Retrieve WAIC.

```{r warning=FALSE}
modelBYMeco.sim$WAIC
```

Check the convergence of the intercept and covariates NO$_2$ and ICU beds.
What do you observe?

```{r echo=TRUE, eval=TRUE, warning = FALSE, fig.height=7, fig.width=10}
mcmc_trace(modelBYMeco.sim$samples, pars = c("alpha", paste0("beta[", 1:K, "]")))
```

Retrieve summary statistics for the two covariates and interpret (it is easier to interpret on the relative scale):

```{r echo=TRUE, eval=TRUE}
modelBYMeco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ]
```

We can get a nice credible intervals plot as well:

```{r warning=FALSE}
modelBYMeco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ] |>
  as_tibble() |>
  select(Median, `95%CI_low`, `95%CI_upp`) |>
  mutate(
    covariate = factor(c("NO2", "ICU", paste0("IMD", 2:5)), levels = c("NO2", "ICU", paste0("IMD", 2:5)))
  ) -> cov.eff

cov.eff |> head()

cov.eff |>
  ggplot(aes(x = covariate, y = Median)) +
  geom_point() +
  geom_errorbar(aes(x = covariate, ymin = `95%CI_low`, ymax = `95%CI_upp`), width = 0.2) +
  ylim(c(0.5, 2.0)) +
  geom_hline(yintercept = 1, lty = 2, col = "red")
```

The effect by unit increase in the long term exposure to NO$_2$ is `RR.beta1_1NO2`:

```{r}
# as relative risk per 1 unit increase in the long term NO2 exposure
modelBYMeco.sim$summary$all.chains[paste0("RR.beta1_1NO2"), c("Median", "95%CI_low", "95%CI_upp")]

# as percentage increase in mortality for 1 unit increase in the long term NO2 exposure
(modelBYMeco.sim$summary$all.chains[paste0("RR.beta1_1NO2"), c("Median", "95%CI_low", "95%CI_upp")] - 1) * 100
```

## 2. Piecewise linear exposure-response regression

We now include a piecewise linear effect for the exposure.
We do this through the covariate effects $\beta X_{i}$.
Specifically, the first two covariates, governed by $\beta_1$ and $\beta_2$, now represent the increase in mortality for low and high concentrations of NO$_2$.
We do this through the design of the covariate matrix $X$: we create an indicator variable $X_{1i}$ which is 1 below average NO$_2$ values and 0 above, and an indicator variable $X_{2i}$ which is 0 below average NO$_2$ values and 1 above.

```{r eval=TRUE, echo = TRUE}
BYMecoPWCode <- nimbleCode({
  # priors
  alpha ~ dflat() # vague prior (Unif(-inf, +inf))
  overallRR <- exp(alpha) # overall RR across study region

  tau.theta ~ dgamma(0.5, 0.05) # prior for the precision hyperparameter
  sigma2.theta <- 1 / tau.theta # variance of unstructured area random effects

  tau.phi ~ dgamma(0.5, 0.0005) # prior on precison of spatial area random effects
  sigma2.phi <- 1 / tau.phi # conditional variance of spatial area random effects

  for (j in 1:K) {
    beta[j] ~ dnorm(0, tau = 1)
    RR.beta[j] <- exp(beta[j])
  }

  # ICAR prior
  phi[1:N] ~ dcar_normal(adj = adj[1:L], weights = weights[1:L], num = num[1:N], tau = tau.phi, zero_mean = 1)

  RR.beta1_1NO2_low <- exp(beta[1] / sd.no2) # get per 1 unit increase in the air pollution for low concentrations (scale back)
  RR.beta1_1NO2_high <- exp(beta[2] / sd.no2) # get per 1 unit increase in the air pollution for high concentrations (scale back)

  # likelihood
  for (i in 1:N) {
    O[i] ~ dpois(mu[i]) # Poisson likelihood for observed counts
    log(mu[i]) <- log(E[i]) + alpha + theta[i] + phi[i] + inprod(beta[], X[i, ])

    SMR[i] <- alpha + theta[i] + phi[i] + inprod(beta[], X[i, ])
    theta[i] ~ dnorm(0, tau = tau.theta) # area-specific RE
    resRR[i] <- exp(theta[i] + phi[i]) # area-specific residual RR
    proba.resRR[i] <- step(resRR[i] - 1) # Posterior probability
  }
})
```

Create data object as required for `NIMBLE`, this time incorporating a piecewise structure for $NO_2$ using indicator variables.

```{r eval=TRUE}
n.LTLA <- dim(data_england)[1]

# create the dummy columns for deprivation
data_england <- data_england |>
  mutate(IMD = as_factor(IMD)) |>
  mutate(as.data.frame(model.matrix(~ 0 + IMD, data = pick(everything()))))

NO2_low = ifelse(scale(data_england$NO2)[, 1] > 0,0,scale(data_england$NO2)[, 1])
NO2_high = ifelse(scale(data_england$NO2)[, 1] < 0,0,scale(data_england$NO2)[, 1])

# matrix of covariates
Xmat <- cbind(
  NO2_low,
  NO2_high,
  scale(data_england$TtlICUB)[, 1],
  data_england$IMD2,
  data_england$IMD3,
  data_england$IMD4,
  data_england$IMD5
)

# Format the data for NIMBLE in a list
covid_data <- list(
  O = data_england$deaths, # observed nb of deaths
  X = Xmat # covariates
)

# number of total covariates
K <- ncol(Xmat)

covid_constants <- list(
  N = n.LTLA, # nb of LTLAs
  K = K, # number of covariates
  sd.no2 = sd(data_england$NO2), # standard deviation of NO2

  # adjacency matrix
  L = length(nbWB_A$weights), # the number of neighboring areas
  E = data_england$expectd, # expected number of deaths
  adj = nbWB_A$adj, # the elements of the neighbouring matrix
  num = nbWB_A$num,
  weights = nbWB_A$weights
)
```

Create the initial values for ALL the unknown parameters:

```{r}
# initialise the unknown parameters, 2 chains
inits <- list(
  list(
    alpha = 0.01,
    beta = rep(0, K),
    tau.theta = 10,
    tau.phi = 1,
    theta = rep(0.01, times = n.LTLA),
    phi = c(rep(0.5, times = n.LTLA))
  ),
  list(
    alpha = 0.5,
    beta = rep(-1, K),
    tau.theta = 1,
    tau.phi = 0.1,
    theta = rep(0.05, times = n.LTLA),
    phi = c(rep(-0.05, times = n.LTLA))
  )
)
```

Which model parameters do you want to monitor?
Set these before running `NIMBLE`.
Call this object `parameters_to_monitor`.

```{r}
parameters_to_monitor <- c("sigma2.theta", "sigma2.phi", "overallRR", "theta", "beta", "RR.beta", "resRR", "proba.resRR", "alpha", "RR.beta1_1NO2_low","RR.beta1_1NO2_high")
```

Run the MCMC simulations using the function `nimbleMCMC()`.
If everything is specified reasonably, this needs approximately 5 minutes.

```{r echo=TRUE, eval=FALSE}
tic <- Sys.time()
modelBYMPWeco.sim <- nimbleMCMC(
  code = BYMecoPWCode,
  data = covid_data,
  constants = covid_constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 50000,
  nburnin = 30000,
  thin = 10,
  nchains = 2,
  setSeed = 9,
  progressBar = TRUE,
  samplesAsCodaMCMC = TRUE,
  summary = TRUE,
  WAIC = TRUE
)
toc <- Sys.time()
toc - tic
# saveRDS(modelBYMPWeco.sim, file = "NIMBLE_BYM_PWLINEAR")
```

```{r}
modelBYMPWeco.sim <- read_rds(here("labs", "exposure_response", "NIMBLE_BYM_PWLINEAR"))
```

Retrieve WAIC and compare with previous model.
Which model performs best?

```{r warning=FALSE}
modelBYMPWeco.sim$WAIC
```

Check the convergence of the intercept and covariates NO$_2$ and ICU beds.
What do you observe?

```{r echo=TRUE, eval=TRUE, warning = FALSE, fig.height=7, fig.width=10}
mcmc_trace(modelBYMPWeco.sim$samples, pars = c("alpha", paste0("beta[", 1:K, "]")))
```

Retrieve summary statistics for the two covariates and interpret (it is easier to interpret on the relative scale):

```{r echo=TRUE, eval=TRUE}
modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ]
```

We can get a nice credible intervals plot as well:

```{r warning=FALSE}
modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ] |>
  as_tibble() |>
  select(Median, `95%CI_low`, `95%CI_upp`) |>
  mutate(
    covariate = factor(c("NO2_LOW", "NO2_HIGH", "ICU", paste0("IMD", 2:5)), levels = c("NO2_LOW", "NO2_HIGH", "ICU", paste0("IMD", 2:5)))
  ) -> cov.eff

cov.eff |> head()

cov.eff |>
  ggplot(aes(x = covariate, y = Median)) +
  geom_point() +
  geom_errorbar(aes(x = covariate, ymin = `95%CI_low`, ymax = `95%CI_upp`), width = 0.2) +
  ylim(c(0.5, 2.0)) +
  geom_hline(yintercept = 1, lty = 2, col = "red")
```

The effect by unit increase in the low concentration long term exposure to NO$_2$ is `RR.beta1_1NO2_low`:

```{r}
# as relative risk per 1 unit increase in the long term NO2 exposure low NO2
modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta1_1NO2_low"), c("Median", "95%CI_low", "95%CI_upp")]

# as percentage increase in mortality for 1 unit increase in the long term NO2 exposure low N02
(modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta1_1NO2_low"), c("Median", "95%CI_low", "95%CI_upp")] - 1) * 100
```

The effect by unit increase in the high concentration long term exposure to NO$_2$ is `RR.beta1_1NO2_high`:

```{r}
# as relative risk per 1 unit increase in the long term NO2 exposure high NO2
modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta1_1NO2_high"), c("Median", "95%CI_low", "95%CI_upp")]

# as percentage increase in mortality for 1 unit increase in the long term NO2 exposure high N02
(modelBYMPWeco.sim$summary$all.chains[paste0("RR.beta1_1NO2_high"), c("Median", "95%CI_low", "95%CI_upp")] - 1) * 100
```

## 3. Non-linear exposure-response regression

Linearity of an exposure response relationship is an assumption. It is also worth exploring whether an exposure-response relationship might be non-linear. Here, we will demonstrate that with a random walk of order 2, which is equivalent to a natural spline with equally-spaced knots.

We create the random walk effect using a ICAR prior with pre-specified weights, which are equivalent models.
We do this in the `weightsRW` function below.

```{r eval=TRUE, echo = TRUE}
BYMecoRW2Code <- nimbleCode({
  # priors
  alpha ~ dflat() # vague prior (Unif(-inf, +inf))
  overallRR <- exp(alpha) # overall RR across study region

  tau.theta ~ dgamma(0.5, 0.05) # prior for the precision hyperparameter
  sigma2.theta <- 1 / tau.theta # variance of unstructured area random effects

  tau.phi ~ dgamma(0.5, 0.0005) # prior on precision of spatial area random effects
  sigma2.phi <- 1 / tau.phi # conditional variance of spatial area random effects

  tau.b ~ dgamma(0.5, 0.0005) # prior on precision of random walk of order 2
  sigma2.b <- 1 / tau.b # conditional variance of random walk of order 2

  # ICAR prior
  phi[1:N] ~ dcar_normal(adj = adj[1:L], weights = weights[1:L], num = num[1:N], tau = tau.phi, zero_mean = 1)

  # random walk prior
  b[1:Jb] ~ dcar_normal(adj = adjb[1:Lb], weights = weightsb[1:Lb], num = numb[1:Jb], tau = tau.b, c = 2, zero_mean = 1)

  for (j in 1:K) {
    beta[j] ~ dnorm(0, tau = 1)
    RR.beta[j] <- exp(beta[j])
  }

  # likelihood
  for (i in 1:N) {
    O[i] ~ dpois(mu[i]) # Poisson likelihood for observed counts
    log(mu[i]) <- log(E[i]) + alpha + theta[i] + phi[i] + inprod(beta[], X[i, ]) + b[Y[i]]
    # the inprod is equivalent to beta[1]*X1[i] + beta[2]*X2[i] + beta[3]*X32[i] + beta[4]*X33[i] + beta[5]*X34[i] + beta[6]*X35[i]

    SMR[i] <- alpha + theta[i] + phi[i] + inprod(beta[], X[i, ]) + b[Y[i]]
    theta[i] ~ dnorm(0, tau = tau.theta) # area-specific RE
    resRR[i] <- exp(theta[i] + phi[i]) # area-specific residual RR
    proba.resRR[i] <- step(resRR[i] - 1) # Posterior probability
  }
})
```

Define the weights of a random walk of order 2:

```{r}
weightsRW <- function(Q) {
  rest.comp <- list()
  for (i in 3:(Q - 2)) {
    rest.comp[[i]] <- c(i - 2, i - 1, i + 1, i + 2)
  }

  rest.comp <- unlist(rest.comp)

  adj <- c(
    2, 3, 1, 3, 4,
    rest.comp,
    c(Q - 3, Q - 2, Q, Q - 2, Q - 1)
  )

  num <- c(2, 3, rep(4, times = c(Q - 4)), 3, 2)

  weights <- c(
    c(2, -1, 2, 4, -1),
    rep(c(-1, 4, 4, -1), times = c(Q - 4)),
    c(-1, 4, 2, -1, 2)
  )

  return(list(adj = adj, num = num, weights = weights))
}
```

Create data object as required for `NIMBLE`, this time processing the $NO_2$ data to be used in a non-linear model:

```{r eval=TRUE}
n.LTLA <- dim(data_england)[1]

# create the dummy columns for deprivation
data_england <- data_england |>
  mutate(IMD = as_factor(IMD)) |>
  mutate(as.data.frame(model.matrix(~ 0 + IMD, data = pick(everything())))) |>
  mutate( # categories for the exposure
    N02_id = as.numeric(cut(NO2, breaks = 100))
  )

Jb <- data_england$N02_id |> max()
W_weights <- weightsRW(Q = Jb)

# matrix of covariates
Xmat <- cbind(
  scale(data_england$TtlICUB)[, 1],
  data_england$IMD2,
  data_england$IMD3,
  data_england$IMD4,
  data_england$IMD5
)

# Format the data for NIMBLE in a list
covid_data <- list(
  O = data_england$deaths, # observed nb of deaths
  X = Xmat # covariates
)

# number of total covariates
K <- ncol(Xmat)

covid_constants <- list(
  N = n.LTLA, # nb of LTLAs
  K = K, # number of covariates

  # adjacency matrix
  L = length(nbWB_A$weights), # the number of neighboring areas
  E = data_england$expectd, # expected number of deaths
  adj = nbWB_A$adj, # the elements of the neighbouring matrix
  num = nbWB_A$num,
  weights = nbWB_A$weights,

  # NO2 categories for random walk
  Y = data_england$N02_id,

  # random walk
  Jb = Jb,
  Lb = length(W_weights$weights),
  adjb = W_weights$adj,
  weightsb = W_weights$weights,
  numb = W_weights$num
)
```

Create the initial values for ALL the unknown parameters:

```{r}
# initialise the unknown parameters, 2 chains
inits <- list(
  list(
    alpha = 0.01,
    beta = rep(0, K),
    tau.theta = 10,
    tau.phi = 1,
    theta = rep(0.01, times = n.LTLA),
    phi = c(rep(0.5, times = n.LTLA)),
    b = rep(0, times = Jb),
    sigma2.b = 0.1
  ),
  list(
    alpha = 0.5,
    beta = rep(-1, K),
    tau.theta = 1,
    tau.phi = 0.1,
    theta = rep(0.05, times = n.LTLA),
    phi = c(rep(-0.05, times = n.LTLA)),
    b = rep(0.5, times = Jb),
    sigma2.b = 1
  )
)
```

Which model parameters do you want to monitor?
Set these before running `NIMBLE`.
Call this object `parameters_to_monitor`.

```{r}
parameters_to_monitor <- c("sigma2.theta", "sigma2.phi", "overallRR", "theta", "beta", "RR.beta", "resRR", "proba.resRR", "alpha", "b", "sigma2.b")
```

Run the MCMC simulations using the function `nimbleMCMC()`.
If everything is specified reasonably, this needs approximately 5 minutes.

```{r echo=TRUE, eval=FALSE}
tic <- Sys.time()
modelBYMRW2eco.sim <- nimbleMCMC(
  code = BYMecoRW2Code,
  data = covid_data,
  constants = covid_constants,
  inits = inits,
  monitors = parameters_to_monitor,
  niter = 50000,
  nburnin = 30000,
  thin = 10,
  nchains = 2,
  setSeed = 9,
  progressBar = TRUE,
  samplesAsCodaMCMC = TRUE,
  summary = TRUE,
  WAIC = TRUE
)
toc <- Sys.time()
toc - tic
# saveRDS(modelBYMRW2eco.sim, file = "NIMBLE_BYM_RW2")
```

```{r}
modelBYMRW2eco.sim <- read_rds(here("labs", "exposure_response", "NIMBLE_BYM_RW2"))
```

Retrieve WAIC and compare with previous model.
Which model performs best?

```{r warning=FALSE}
modelBYMRW2eco.sim$WAIC
```

Check the convergence of the intercept and covariates NO$_2$ and ICU beds.
What do you observe?

```{r echo=TRUE, eval=TRUE, warning = FALSE, fig.height=7, fig.width=10}
mcmc_trace(modelBYMRW2eco.sim$samples, pars = c("alpha", paste0("beta[", 1:K, "]")))
```

Retrieve summary statistics for the two covariates and interpret (it is easier to interpret on the relative scale):

```{r echo=TRUE, eval=TRUE}
modelBYMRW2eco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ]
```

We can get a nice credible intervals plot as well:

```{r warning=FALSE}
modelBYMRW2eco.sim$summary$all.chains[paste0("RR.beta[", 1:K, "]"), ] |>
  as_tibble() |>
  select(Median, `95%CI_low`, `95%CI_upp`) |>
  mutate(
    covariate = factor(c("ICU", paste0("IMD", 2:5)), levels = c("ICU", paste0("IMD", 2:5)))
  ) -> cov.eff

cov.eff |> head()

cov.eff |>
  ggplot(aes(x = covariate, y = Median)) +
  geom_point() +
  geom_errorbar(aes(x = covariate, ymin = `95%CI_low`, ymax = `95%CI_upp`), width = 0.2) +
  ylim(c(0.5, 2.0)) +
  geom_hline(yintercept = 1, lty = 2, col = "red")
```

Plot the non-linear $NO_2$ association
```{r warning=FALSE}
# logic to get the first value of the bin
NO2_bins <- data_england |>
  pull(NO2) |>
  cut(100) |>
  str_sub(2, -2) %>%
  sub("\\,.*", "", .) |>
  as.numeric() |>
  unique() |>
  sort()

# helper function to convert nimble output to one favoured by posterior and bayesplot
as_mcmc_obj <- \(x) as_draws_array(as.mcmc.list(lapply(x, as.mcmc)))

exposure_posterior <- as_mcmc_obj(modelBYMRW2eco.sim$samples)

##
## Define the x-axis of NO2
X_axis <- cut(data_england$NO2, breaks = 100)
X_axis <- X_axis %>% levels()
X_axis
get_midpoint <- function(cut_label) {
  mean(as.numeric(unlist(strsplit(gsub("\\(|\\)|\\[|\\]", "", as.character(cut_label)), ","))))
}
get_midpoint(X_axis)
sapply(X_axis, get_midpoint) %>% as.numeric() -> X_axis

##
## and plot
summarise_draws(exposure_posterior, default_summary_measures()) |>
  filter(startsWith(variable, "b[")) |>
  mutate(X = X_axis) |>
  ggplot() +
  geom_line(aes(x = X, y = median)) +
  geom_ribbon(aes(x = X, ymin = q5, ymax = q95), fill = "grey80", alpha = 0.5) +
  labs(x = "NO2", y = "log mortality risk")
```

We can also show the effect on the relative scale, relative to the risk at 0 concentration:

```{r warning=FALSE}
exposure_posterior[, , str_c("b[", 1:100, "]")] |>
  exp() %>%
  sweep(x = ., MARGIN = c(1, 2), FUN = "/", STATS = exp(extract_variable_matrix(exposure_posterior, "b[50]"))) %>%
  apply(X = ., MARGIN = 3, quantile, probs = c(0.5, 0.025, 0.975)) |>
  t() |>
  as_tibble() |>
  mutate(x = X_axis) |>
  ggplot() +
  geom_line(aes(x = x, y = `50%`)) +
  geom_ribbon(aes(x = x, ymin = `2.5%`, ymax = `97.5%`), fill = "grey80", alpha = 0.5) +
  labs(x = "NO2", y = "Relative mortality risk") 
```

## Closing remarks

Here, we examined the effect of long term exposure to NO$_2$ on COVID-19 mortality.
We fitted several models, with either linear or non-linear models.
We fitted a BYM model to account for unknown spatial confounding but in addition we accounted for total number of ICU beds and deprivation per LTLA.
We reported evidence of an increased COVID-19 mortality for increasing levels of NO$_2$.

It should be noted that this model is not fully adjusted, and that the actual analysis from the paper was at a higher resolution and with full adjustment by potential confounders via covariates. The data from that paper is sensitive and cannot be shared publicly. Nevertheless, this lab shows the basic mechanics of how to create linear and non-linear exposure-response analyses in a Bayesian framework. 
